---
title: "TREXSeg - cell segmentation workflow"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation
***

To install TREXSeg and run this tutorial you should first make sure to install EBImage from Bioconductor. You can then isntall the package from GitHub using the devtools R package.

```{r install, eval=FALSE}

# Install EBImage
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("EBImage")

# Install TREXSeg
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("devtools")

devtools::install_github("ludvigla/TREXSeg")

```

```{r load_libs, warning=FALSE, message=FALSE}

library(EBImage)
library(TREXSeg)

```

## Load data
***

To demonstrate the segmentation approach, we will look an immunofluorscence image of the neuronal marker NeuN. 

First, let's crop out a smaller rectangle of the image to get a finer detail of the cells. The images are stored as 16bit TIFFs so when loading the data we can normalize the intensity values so that the highest intensity values are close to 1. In the plot you can see all the stained nuclei but it's quite difficult to distinguish them from autofluorescent speckles and background. 

```{r load_example_image, warning=FALSE, fig.width=6, fig.height=6}

neun <- system.file("data", paste0("V13/neun.tif"), package = "TREXSeg")
neun.im <- readImage(files = neun)
neun.im <- EBImage::normalize(neun.im)

cat("Image dimensions: ", dim(neun.im))

# Crop
neun.im <- neun.im[1500:2000, 1500:2000]
plot(neun.im)
title(main = "Raw", col.main = "cyan")

```

## 2D FFT filter
***

Next we can apply a 2D convolution to blur the images and even out small artifacts. The difference is quite subtle, but you should see that the cells are a bit blurrier than in the previous image.

```{r filter_example_image4, warning=FALSE, fig.width=12, fig.height=6, out.width = "100%"}

neun.im.filtered <- filter_cells(neun.im)

par(mfrow = c(1, 2), mar = c(0, 0, 0, 0))
plot(neun.im)
box(col = "cyan")
title(main = "Raw", col.main = "cyan", line = -2)
plot(neun.im.filtered)
title(main = "Filtered", col.main = "cyan", line = -2)
box(col = "cyan")

```

## Correction
***

Using the raw image and the filtered image, we can now subtract the blurry image from the raw image which makes the cells stand out much clearer than before.

```{r filter_example_image2, warning=FALSE, fig.width=12, fig.height=6, out.width = "100%"}

neun.im.corrected <- correct_cells(im = neun.im, imfiltered = neun.im.filtered)

par(mfrow = c(1, 2), mar = c(0, 0, 0, 0))
plot(neun.im)
title(main = "Raw", col.main = "cyan", line = -2)
box(col = "cyan")
plot(neun.im.corrected)
title(main = "Corrected", col.main = "cyan", line = -2)
box(col = "cyan")

```

## Thresholding
***

Now that the cells are clearly distinguishable from the background, we can apply a threshold to the corrected image to segment out the cells nuclei. The thresholding method is a very simple mean + standard deviation threshold, i.e. pixels with an intensity value higher than mean(intensity) + nsd*sd(intensity) are considered to come from cell nuclei. The default value for nsd is 2, meaning that the threshold is set 2 standard deviations above the mean. Setting a higher value of nsd will result in fewer nuclei pixels and most of the time also fewer and/or smaller cells.

```{r filter_example_image3, warning=FALSE, fig.width=12, fig.height=6, out.width="100%"}

neun.im.thresh <- threshold_cells(imcorrected = neun.im.corrected)

par(mfrow = c(1, 2), mar = c(0, 0, 0, 0))
plot(neun.im)
title(main = "Raw", col.main = "cyan", line = -2)
box(col = "cyan")
plot(neun.im.thresh)
title(main = "Thresholded", col.main = "cyan", line = -2)
box(col = "cyan")

```

## Cleaning
***

You can see that there are quite a lot of small speckles left in the image (and sometimes very big shapes as well) which we can remove by cleaning out objects with a predefined area range. If we plot a histgram of shape areas we can select a lower and upper bound for the allowed shape sizes. 

```{r histogram}

hist(table(bwlabel(neun.im.thresh))[-1], breaks = 30, main = "Histogram of shape areas", xlab = "pixel area")

```

```{r clean_example_image, warning=FALSE, fig.width=12, fig.height=6}

neun.im.clean <- clean_cells(imthreshold = neun.im.thresh)

par(mfrow = c(1, 2), mar = c(0, 0, 0, 0))
plot(neun.im)
title(main = "Raw", col.main = "cyan", line = -2)
box(col = "cyan")
plot(neun.im.clean)
title(main = "Cleaned", col.main = "cyan", line = -2)
box(col = "cyan")

```

## Watershed
***

Sometimes multiple nuclei are merged together because they are closely packed. The watershed algorithm makes it possible to separate some of these merged nuclei. After running the watershed step, we have segmented cells each and each cell is labelled with a unique ID. Note that if you run `bwlabel`, you will effect of the watershedding and change the cell labels.

```{r clean_example_image2, warning=FALSE, fig.width=12, fig.height=6, out.width="100%"}

neun.im.watershed <- watershed_cells(imclean = neun.im.clean)

par(mfrow = c(1, 2), mar = c(0, 0, 0, 0))
plot(neun.im)
box(col = "cyan")
title(main = "Raw", col.main = "cyan", line = -2)
plot(colorLabels(neun.im.watershed))
box(col = "cyan")
title(main = "Labelled", col.main = "cyan", line = -2)

```

# Data
***

We can now run the whole workflow using the `SegmentCells` function. To demonstrate the workflow we have a test dataset including Immunofluoresence imaging of Egfp, Neun and Olig2 for a total of four tissue sections. 


```{r segmentation_run}

# list input files
channels <- c("egfp", "neun", "olig2")
img.files <- setNames(system.file("data", paste0("V13/", channels, ".tif"), package = "TREXSeg"), nm = channels)

```

The threshold ranges used for the cleaning step can be adjusted for each target. You can also use a predefine "crop window" if you want to run the segmentation on a smaller part of the image.

NOTE: All images used here have a size of approximatly 4400*4400 pixels. If you run the segmentation on higher resolution images you should adjust the filters and to values that are appropriate for the nuclei sizes.

```{r read_segmented, include=FALSE}
segmented.list <- readRDS(system.file("data/RData/segmented.list", package = "TREXSeg"))
```
```{r run segmentation, warning=FALSE, eval=FALSE}

# define feature threshold ranges
feature.thresholds <- list(neun = c(5, 40), 
                           olig2 = c(8, 40), 
                           egfp = c(5, 40))

# read images and run segmentation for each channel
segmented.list <- setNames(lapply(seq_along(img.files), function(i) {
  target <- names(img.files)[i]
  SegmentCells(
    impath = img.files[i], 
    crop.window = c(3000, 3500, 2500, 3000),
    feature.threshold = feature.thresholds[[target]], 
    verbose = TRUE)
}), nm = c("egfp", "neun", "olig2"))

```

Now we can plot the segmented nuclei for any of the four sections. We can for example encode olig2, egfp and neun are RGB colors to visualize them simultaneously. 

Below the oligodendocytes are colored red, Egfp+ nuclei green and Neurons blue. There are also a few nuclei which are olig2+Egfp+ (yellow) or NeuN+Egfp+ (cyan).

```{r plot_seg, fig.width=6, fig.height=6}

im <- rgbImage(red = segmented.list$olig2, 
               green = segmented.list$egfp,
               blue = segmented.list$neun)
plot(im)
title(main = "red : Oligodendrocytes, green : Egfp, blue : Neurons", col.main = "cyan")

```

## Count detected nuclei
***

Now we have segmented nuclei Egfp, Neun and Olig2 from four tissue sections. We can count the number of detected nuclei using the `table` function. Applying `table` on the segmented output will count the number of pixels of each set of "connected" pixels, i.e. cells which is equal to the pixel area. We remove the first element of the ouput from `table` which is the area of the background.

```{r count_features}

for (s in names(segmented.list)) {
  segmented <- segmented.list[[s]]
  cat("Total number of oligodendrocytes in ", s, ": ", length(table(segmented.list$olig2[-1])), "\n")
  cat("Total number of EGFP in ", s, ": ", length(table(segmented.list$egfp[-1])), "\n")
  cat("Total number of neurons in ", s, ": ", length(table(segmented.list$neun[-1])), "\n\n")
}

```

## Find overlapping cells
***

If we want to detect nuclei with overlapping signals from two channels, we can run `OverlapImages`. 

```{r overlap_computation, fig.width=6, fig.height=6}

im <- rgbImage(green = segmented.list$egfp, 
               blue = segmented.list$neun)
par(mar = c(0, 0, 0, 0))
plot(im)

olig2_neun <- OverlapImages(segmented.list$egfp, segmented.list$neun) 
title(main = "green : Egfp, blue : Neurons, cyan : Egfp+ Neurons", col.main = "cyan", line = -2)

```

The `OverlapImages` function makes sure that only nuclei with a minimum predefined overlap (at least 50% of the smallest area) are kept. 

```{r plot_overlap, fig.width=12, fig.height=4}

par(mfrow = c(1, 3), mar = c(0, 0, 0, 0))
plot(segmented.list$egfp)
box(col = "cyan")
title(main = "Egfp", col.main = "cyan", line = -2)
plot(segmented.list$neun)
box(col = "cyan")
title(main = "NeuN", col.main = "cyan", line = -2)
plot(olig2_neun)
box(col = "cyan")
title(main = "Overlapping Egfp and Neun", col.main = "cyan", line = -2)

```

## Count Egfp+ nuclei
***

We can extract different features from the segmented nuclei using the `computeFeatures` methods from EBImage.

```{r estimate_counts, fig.width=8, fig.height=4}

par(mfrow = c(1, 2))

# x, y coordinates
xy.coords <- computeFeatures.moment(x = olig2_neun)
plot(xy.coords[, c("m.cx", "m.cy")], pch = 19)
title(main = "nuclei positions")

# shape features
shape.fts <- computeFeatures.shape(x = olig2_neun)
boxplot(shape.fts[, "s.area"])
title(main = "nuclei areas [number of pixels]")

```


